Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-05-08 VegOzone.tex
Tune text in Section 2

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on conditions during the 2018/19 growing season at the NIBIO operated Environment Centre Svanhovd. To this end, we collected ozone and atmospheric monitoring data taken at Svanhovd from different distributors. The main reasons for this spread are varring update frequencies and length of available data sets. In terms of ozone, we also use observational data from monitoring sites in northern Finland and Sweden for cross-evaluation. Svanhovd synonymously refers to the atmospheric monitoring site at Svanvik/Pasvik throughout the paper.

%\subsection{NIBIO Environment Centre Svanhovd and bio-monitoring}
%\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

To relate observed visible damage on ozone sensitive plants to local ozone concentration (\chem{[O_3]}), both an ozone bio-monitor (referred to as ozone garden) and a conventional ozone monitor using the existing atmospheric monitoring infrastructure had been installed during the 2018/19 growing season at the NIBIO centre. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik, where the Pasvik river, which is the border to Russia, lies only a few hundred meters to the east. This ozone garden was likely the northern-most to date {\bf TODO: Reference?}. An ozone garden consists of selected plant species that are very sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed qualitatively visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

All measured atmospheric variables including ozone for the 2018/19 growing season were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. The ozone monitoring data have been added to the EBAS database operated by \citet{NILU_EBAS} and are thus openly accessible. All long-term ozone observation data described in Section~\ref{subsec:ozone_data} are also obtained from EBAS. The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

@@

As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak \chem{[O_3]}. In contrast, \chem{[O_3]} only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019. For further analysis (Section~\ref{sec:stats}) and modeling study with the $\mathrm{DO_3SE}$ model (Section~\ref{sec:do3se}), we reconstructed the missing data. The used gap-filling method will be described in detail in a companion paper to be submitted.

@@

\subsection{Ozone data}
\label{subsec:ozone_data}
To determine the significance of the higher \chem{[O_3]} in 2018 in relation to regional averages and trends, we derive a ground-level \chem{[O_3]} climatology for Svanvik (Section~\ref{sec:stats}). From 1986--1996, ozone monitoring has been operated continuously by NILU at Svanhovd, but no ozone measurements have been conducted before the 2018/19 growing season. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we cross-calibrate our data with long-term ozone observations at sites in northern Finland and Sweden.

@@

Data from before the 1990s are available solely from Svanvik and Karasjok. These data, however, do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably leads to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream of the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a breaking point in the combined data series, denoted as Jergul/Karasjok in the following, which would be particularly problematic if these were the sole data for comparison. Only a few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are shown for completeness. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas indicate periods with insufficient quality control as mentioned above.

@@

\section{Statistical analysis}
\label{sec:stats}
In this section, we assess the climate conditions and multi-annual mean (referred to as climatology in the following) annual cycle of gound-level ozone concentrations at Svanhovd. To this end, we derive climatologies for temperature, precipitation, global irradiance, and ozone. We evaluate statistical significance of divergances from the norm in these variables for the growing seasons 2018/19 (referred to as anomalies). We perform a bias correction on the Svanvik ozone climatology with respect to long-term series of \chem{[O_3]} at the sites in northern Finland and Sweden.

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth which is shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from the atmospheric monitoring in Svanvik (1992--2012). Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with a standard error of the mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classification of a regular subarctic climate (Dfc) \citep{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{climatediagram}
  \includegraphics[width=8.3cm]{fig05}
  \caption{Climate diagram following Walter and Lieth. The diagram is based on climatological data for Svanvik/Pasvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged temperatures (red line) are displayed with standard error of the mean as error band. Averaged monthly accumulated precipitation (blue bars) is shown with a standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star.}
  \label{fig:climatediagram}
\end{figure}

@@

The derived climatology for global irradiance ($Q_0$) at Svanvik is shown in Appendix Fig.~\ref{fig:global_rad_clim}a). Shown are the average of hourly maximum $\left<Q_0^\mathrm{max}\right>$, minimum $\left<Q_0^\mathrm{min}\right>$, and mean $\left<Q_0^\mathrm{mean}\right>$. In agreement with the geometrically derived sunshine duration displayed in Supplement Fig.~\ref{fig:sunlight_fennoscandia}, $\left<Q_0^\mathrm{max}\right>$ approaches zero in December/January commonly known as polar night and reaches its maximum in June/July commonly referred to as midnight sun conditions. The highest observed $\left<Q_0^\mathrm{max}\right>$ is about $800\,\unit{W\,m^{-2}}$, while on average $\left<Q_0^\mathrm{mean}\right>$ it is about $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the climatological minimum $\left<Q_0^\mathrm{min}\right>$ lies at about $200\,\unit{W\,m^{-2}}$.

@@

The \chem{\left<\Delta\left<[O_3]\right>\right>} for Svanvik in 2018 is slightly larger than for Esrange and Pallas. This could mean either that 2018 had been more exceptional at Svanvik or the derived climatology may need a correction for increased tropospheric background ozone concentration. Based on the meteorological conditions, we assume 2019 as an average year which leads to an estimated constant bias of about $1.2\,\unit{ppb}$ in \chem{[O_3]}. Daily \chem{\Delta_{present-past}\left<[O_3]\right>} from Esrange and Pallas, wherein present is defined as the average of 2010--2018 and past as the respective beginning of data taking until 1997, do not reveal a clear trend: We find for Esrange $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas $(-2.5\pm 3.7)\,\unit{ppb}$ \citep{NILU2003, BER:Ruoho-Airola2015}. Former is compatible with our assumed bias for the climatology derived for Svanvik.

>>>>

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on conditions during the 2018/19 growing season at the NIBIO operated Environment Centre Svanhovd. We also set these into a climatological and regional context. To this end, we collected atmospheric and ozone monitoring data taken at Svanhovd and additional ozone monitoring data from observation sites in northern Finland and Sweden for cross-evaluation. Varring update frequencies and length of available data sets made it necessary to fetch the data from various sources (further details below). Svanhovd synonymously refers to the atmospheric monitoring site Svanvik/Pasvik throughout the paper.\\

The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The NIBIO centre Svanhovd is located south of the settlement of Svanvik, where the Pasvik river, which is the border to Russia, lies only a few hundred meters to the east. It focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

To relate local ozone concentration (\chem{[O_3]}) with visible damage on ozone sensitive plants, both an ozone bio-monitor (referred to as ozone garden) and a conventional ozone monitor using the existing atmospheric monitoring infrastructure had been installed during the 2018/19 growing season at the NIBIO centre. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). This ozone garden was likely the northern-most to date {\bf TODO: Reference?}. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed qualitatively visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

All measured atmospheric variables including ozone for the 2018/19 growing season were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. The ozone monitoring data have been also added to the EBAS database operated by \citet{NILU_EBAS} and are thus openly accessible. All long-term ozone observation data described in Section~\ref{subsec:ozone_data} are also obtained from EBAS. The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

@@

As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition only started in late April. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak \chem{[O_3]}. In contrast, \chem{[O_3]} only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019. For further analysis (Section~\ref{sec:stats}) and modeling study with the $\mathrm{DO_3SE}$ model (Section~\ref{sec:do3se}), we reconstructed the missing data. The used gap-filling method will be described in detail in a companion paper to be submitted.

@@

\subsection{Ozone data}
\label{subsec:ozone_data}
To determine the significance of the higher \chem{[O_3]} in 2018 in relation to regional averages and trends, we derive a ground-level \chem{[O_3]} climatology for Svanvik (Section~\ref{sec:stats}). From 1986--1996, ozone monitoring has been operated continuously by NILU at Svanhovd, but no ozone measurements have been conducted before the 2018/19 growing season. Due to probable changes in northern hemispheric tropospheric background \chem{[O_3]} and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we cross-calibrate our data with long-term ozone observations at sites in northern Finland and Sweden.

@@

Data from before the 1990s are available solely from Svanvik and Karasjok. These data, however, do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably leads to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream of the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a baseline shift in the combined data series, denoted as Jergul/Karasjok, which would be particularly problematic if these were the sole data for comparison. Only a few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are listed for completeness. There is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik (Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}).

@@

\section{Statistical analysis}
\label{sec:stats}
In this section, we assess the climate conditions. To this end, we compute multi-annual mean (referred to as climatology) of environmental key variables (temperature, precipitation, global irradiance) and gound-level \chem{[O_3]} at Svanhovd. We evaluate the statistical significance of divergances from the norm in these variables for the growing seasons 2018/19 (referred to as anomalies) and perform a bias correction on the ozone climatology at Svanvik accounting for the increase in tropospheric background \chem{[O_3]}. For this, we cross-calibrate with long-term observations of \chem{[O_3]} at the sites in northern Finland and Sweden.

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram based on climatological data from the atmospheric monitoring in Svanvik (1992--2012) (Fig.~\ref{fig:climatediagram}) supports this very well. Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with a standard error of the mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppen's climate classification of a regular subarctic climate (Dfc) \citep[][e.g.]{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature ($29.4\,\unit{^\circ C}$) occurred July 16 the same year.

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{climatediagram}
  \includegraphics[width=8.3cm]{fig05}
  \caption{Climate diagram following Walter and Lieth. The diagram is based on climatological data for Svanvik/Pasvik (1992--2012). Monthly averaged temperatures (red line) are displayed with standard error of the mean as error band. Averaged monthly accumulated precipitation (blue bars) is shown with a standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star.}
  \label{fig:climatediagram}
\end{figure}

@@

The derived climatology for global irradiance ($Q_0$) at Svanvik (Appendix Fig.~\ref{fig:global_rad_clim}a)) in agreement with the geometrically derived sunshine duration (Supplement Fig.~\ref{fig:sunlight_fennoscandia}), approaches zero in December/January (polar night) and reaches its maximum in June/July (midnight sun conditions). The highest observed averaged daily maximum $\left<Q_0^\mathrm{max}\right>$ is about $800\,\unit{W\,m^{-2}}$, while on averaged daily mean $\left<Q_0^\mathrm{mean}\right>$ it is about $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the averaged daily minimum $\left<Q_0^\mathrm{min}\right>$ lies at about $200\,\unit{W\,m^{-2}}$.

@@

The \chem{\left<\Delta\left<[O_3]\right>\right>} for Svanvik in 2018 is slightly larger than for Esrange and Pallas. This could mean either that 2018 had been more exceptional at Svanvik or the derived climatology may need a correction for increased tropospheric background \chem{[O_3]}. Based on the meteorological conditions, we assume 2019 as an average year which leads to an estimated constant bias of about $1.2\,\unit{ppb}$ in \chem{[O_3]}. Daily \chem{\Delta_{present-past}\left<[O_3]\right>} from Esrange and Pallas, wherein present is defined as the average of 2010--2018 and past as the respective beginning of data taking until 1997, do not reveal a clear trend: We find for Esrange $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas $(-2.5\pm 3.7)\,\unit{ppb}$ \citep{NILU2003, BER:Ruoho-Airola2015}. Former is compatible with our assumed bias for the climatology derived for Svanvik.
